---
title: BUSA8090 Assignment 2- Visualisation Task
author: Bea Lynor Barros 47758295
date: "2023-11-01"
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, package, echo=FALSE}

suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))


library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)
library(reshape2)

```

```{r, data, echo=FALSE}

#Data Source Information
# Dataset: Most Streamed Spotify Songs
# Source: BUSA8090 Assignment 2 File


# Load the Spotify dataset
spotify_dataset <- read.csv("spotify-2023.csv", na.strings = c("NaN", "NA"))

# Create a copy of the Spotify dataset
spotify_data <- spotify_dataset


#check column names
colnames(spotify_data)

```

***Data Cleaning and Wrangling***

#rename column names to make it more organized

```{r}
# Clean and rename column names

colnames(spotify_data) <- c(
  "Track_Name", "Artist_Name", "Artist_Count",
  "Released_Year", "Released_Month", "Released_Day",
  "In_Spotify_Playlists", "In_Spotify_Charts", "Streams",
  "In_Apple_Playlists", "In_Apple_Charts", "In_Deezer_Playlists",
  "In_Deezer_Charts", "In_Shazam_Charts", "BPM",
  "Key", "Mode", "Danceability",
  "Valence", "Energy", "Acousticness",
  "Instrumentalness", "Liveness", "Speechiness"
)

head(spotify_data)

```

#Rename typo in streams column
```{r}

# Replace a specific value with a blank (empty string) in the "Streams" column

spotify_data$Streams[spotify_data$Streams == "BPM110KeyAModeMajorDanceability53Valence75Energy69Acousticness7Instrumentalness0Liveness17Speechiness3"] <- ""

```

#convert datatypes
```{r}

#group numeric variables 
numeric_features <- c("Streams", "In_Shazam_Charts", "In_Deezer_Playlists", "Artist_Count", "Released_Year","Released_Month","Released_Day","In_Spotify_Playlists","In_Spotify_Charts","In_Apple_Playlists", "In_Apple_Charts", "In_Deezer_Charts", "BPM")

#group factor variables 
factor_features <- c("Artist_Name", "Key", "Mode", "Track_Name")

#convert accordingly
spotify_data[numeric_features] <- lapply(spotify_data[numeric_features], as.numeric)
spotify_data[factor_features] <- lapply(spotify_data[factor_features], as.factor)

# Check the data types of all columns and print
str(spotify_data)

```


***Visualization 1: Distribution of Song Features***

```{r}

# Pivoting the data
long_spotify_data <- spotify_data %>%
    pivot_longer(
        cols = c("Danceability", "Valence", "Energy", "Acousticness", "Instrumentalness", "Liveness", "Speechiness"), 
        names_to = "Property", 
        values_to = "Value"
    )

# Plotting the data
plot_song_distribution <- ggplot(long_spotify_data, aes(x = Value)) + 
    geom_histogram(binwidth = 0.025, fill = "#0072B2", color = "black") + 
    facet_wrap(~Property, scales = "free", ncol = 2) +
    labs(title = "Distribution of Song Features",
         x = "Song Feature",
         y = "Count of Songs") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 13, face = "bold"),
          strip.background = element_blank(),
          strip.text = element_text(face = "bold"),
          axis.title = element_text(size = 12),
          axis.text = element_text(size = 7))

#show plot
plot_song_distribution


```


***Visualization 2: Heatmap of Song Features***

```{r}
# Specify the selected song features
correlation_variables <- spotify_data[, c("Danceability", "Valence", "Energy", 
                                          "Acousticness", "Instrumentalness", 
                                          "Liveness", "Speechiness", "Released_Year")]

# Create the correlation matrix
corr_spotify <- round(cor(correlation_variables), 2)

# Reshape the correlation matrix for plotting
melted_corr_mat_spotify <- melt(corr_spotify)

# Plot the correlation heatmap
heatmap_songfeature <- ggplot(data = melted_corr_mat_spotify, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 3) +
  labs(title = "Song Feature Heatmap",
       x = "Features",
       y = "Features") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(angle = 0, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))  

# Print the heatmap
heatmap_songfeature

```

***Visualization 3: Boxplot of Song Features***

```{r}

# List of audio features
song_features <- c("Danceability", "Valence", "Energy", "Acousticness", "Instrumentalness", "Liveness", "Speechiness")

# Reshape the data into a long format
long_spotify_data <- spotify_data %>%
  select(all_of(song_features)) %>%
  pivot_longer(cols = everything(), names_to = "Feature", values_to = "Value")

# Create the boxplot
boxplot_songfeature <- ggplot(long_spotify_data, aes(x = Feature, y = Value, fill = Feature)) +
  geom_boxplot() +
  labs(title = "Boxplot of Song Features",
       x = "Song Feature",
       y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))  # Center the title

#print plot
boxplot_songfeature

```

***Visualization 4: Songs per Decade***

```{r}

# Define the breaks for the Released_Year groups
year_breaks <- c(1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020, 2030)

# Define the labels for the groups
year_labels <- c("30s", "40s", "50s", "60s", "70s",
                 "80s", "90s", "2000s", "2010s", "2020s")

# Add a new column "Year_Group" with the group labels
spotify_data <- spotify_data %>%
  mutate(Year_Group = cut(Released_Year, breaks = year_breaks, labels = year_labels, right = FALSE))

# Group by Year_Group and calculate total tracks and percentage share
decade_released_songs <- spotify_data %>%
  group_by(Year_Group) %>%
  summarise(Total_Tracks = n()) %>%
  mutate(Percentage_Share = round(Total_Tracks / sum(Total_Tracks) * 100, 2)) %>%
  arrange(desc(Total_Tracks))

# Print the summary data
print(decade_released_songs)
```


```{r}
# Create the bar chart
plot_tracksperdecade <- ggplot(decade_released_songs, aes(x = Total_Tracks, y = Year_Group, 
                                                 label = paste(Percentage_Share, "%"), 
                                                 fill = Year_Group)) +
  geom_bar(stat = "identity") +
  geom_text(aes(x = Total_Tracks), size = 2.7, vjust = -0.5, hjust = 0.5) +
  labs(title = "Total Tracks per Decade", x = "Total Tracks", y = "Decade") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 0.5), 
        plot.title = element_text(hjust = 0.5, face = "bold"))

#print the plot
plot_tracksperdecade

```

***Viualization 5: Most Streamed artist***

```{r}

#Create table of top 20 most streamed artists
top_20_artists <- spotify_data %>%
  group_by(Artist_Name) %>%
  summarize(TotalStreams = sum(Streams)) %>%
  arrange(desc(TotalStreams)) %>%
  slice(1:20)

#print result
top_20_artists

```

```{r}

# Create the barplot
plot_top_20_artists <- ggplot(top_20_artists, 
                              aes(x = reorder(Artist_Name, -TotalStreams), 
                                  y = TotalStreams, 
                                  label = comma(TotalStreams), 
                                  fill = Artist_Name)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 20 Most Streamed Artists",
       x = "Artist Name",
       y = "Total Streams") +
  scale_y_continuous(labels = comma, expand = c(0.1, 0.1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 7),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        legend.key.size = unit(0.3, "cm"))

# Print the plot
plot_top_20_artists

```

***Viualization 6: Distribution of Keys by Major or Minor***

```{r}

# Exclude rows where Key is blank
spotify_data_filtered <- spotify_data %>%
  filter(!(Key %in% c("")))

# Count the occurrences of "Track_Name" within each "Key & Mode" for the filtered data
spotify_KeyMode_count <- spotify_data_filtered %>%
  group_by(Mode, Key) %>%
  summarise(Count = n())

# Calculate the total count
total_count <- sum(spotify_KeyMode_count$Count)


# View the final dataframe
print(spotify_KeyMode_count)


```

```{r}

# Plot the table
plot_spotify_KeyMode_count <- ggplot(spotify_KeyMode_count, 
                                     aes(x = reorder(Key, -Count), 
                                         y = Count, 
                                         fill = Mode)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Keys by Major or Minor",
       x = "Key",
       y = "Total Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom") +
  geom_text(aes(label = Count), 
            position = position_stack(vjust = 0.5), 
            color = "white", 
            size = 3)


# Print the graph
plot_spotify_KeyMode_count

```
